%----------------------------------------------------------------------------
\chapter{The two algorithms}
\label{chap:Algorithms}
%----------------------------------------------------------------------------

This chapter states a simplified form of the XPBD and VBD algorithms I implemented in
\autoref{chap:Methodology}. Both target the same variational problem \eqref{eq:G},
but they reach it through opposite descent directions.

%----------------------------------------------------------------------------
\section{XPBD: dual-variable position projection}
\label{sec:XPBDAlgo}
%----------------------------------------------------------------------------

\subsection{Derivation}
\label{ssec:XPBDDeriv}
For each elastic spring, XPBD associates a constraint $C(\mathbf{x}) = 0$ at zero
deformation with a compliance $\alpha = 1/k$, the inverse of the spring stiffness.
Because the implicit time stepping (\autoref{ssec:ImplicitEuler}) carries a factor
of $1/h^{2}$ on the inertial term of \eqref{eq:G}, the compliance enters the update
only after being scaled by the squared substep size. Folding that factor into the
compliance defines the \emph{time-step-scaled compliance}
$\tilde{\alpha} = \alpha / h^{2}$, which is what lets XPBD reproduce an effective
stiffness that is independent of $h$ and of the iteration count. The quantity
$\lambda = -\tilde{\alpha}^{-1} C(\mathbf{x})$ is the Lagrange multiplier for that constraint and
can be interpreted as the magnitude of the spring force. The augmented Lagrangian
of \eqref{eq:G} with a single substep $h$ has the stationarity condition
\begin{align}
    \mathbf{M}(\mathbf{x} - \mathbf{y}) - \nabla \mathbf{C}(\mathbf{x})^{T}\,\boldsymbol{\lambda} &= \mathbf{0}, \\
    \mathbf{C}(\mathbf{x}) + \tilde{\alpha}\,\boldsymbol{\lambda} &= \mathbf{0},
\end{align}
where $\mathbf{C}$ and $\boldsymbol{\lambda}$ stack the per-constraint functions and
multipliers. XPBD solves this
coupled system with a Gauss--Seidel sweep over the constraints: for each
constraint $j$, the multiplier increment is
\begin{equation}
    \label{eq:XPBDLambda}
    \Delta \lambda_{j} \;=\; \frac{-C_{j}(\mathbf{x}) - \tilde{\alpha}_{j}\, \lambda_{j}}
                                {\nabla C_{j}\,\mathbf{M}^{-1}\nabla C_{j}^{T}
                                 + \tilde{\alpha}_{j}},
\end{equation}
and the position correction is
\begin{equation}
    \Delta \mathbf{x} = \mathbf{M}^{-1}\nabla C_{j}^{T}\,\Delta \lambda_{j}.
\end{equation}
Velocities are recovered at the end of each substep by
$\mathbf{v}_{t+h} = (\mathbf{x}_{t+h} - \mathbf{x}_{t}) / h$.

\subsection{Algorithm}
\label{ssec:XPBDPseudo}
Algorithm \autoref{alg:XPBDCore} shows the simplified form of XPBD used for the core
comparison: no damping, collision handling.
The full implemented version is given in \autoref{sec:AppXPBD} with a more detailed pseudocode.

\begin{algorithm}[!ht]
\caption{XPBD step (core comparison). $n_{\mathrm{sub}}$ substeps, each with
         $n_{\mathrm{it}}$ Gauss--Seidel iterations over the constraints.}
\label{alg:XPBDCore}
\begin{algorithmic}[1]
\State $h \gets dt / n_{\mathrm{sub}}$ \Comment{substep size}
\For{$s = 1$ \textbf{to} $n_{\mathrm{sub}}$}
    \ForAll{vertices $i$}
        \State $\mathbf{v}_i \gets \mathbf{v}_i + h\bigl(\mathbf{g} + w_i\,\mathbf{f}_{\mathrm{ext}}(\mathbf{x}_i)\bigr)$
               \Comment{$w_i = 1/m_i$}
        \State $\mathbf{x}^{\mathrm{prev}}_i \gets \mathbf{x}_i$
        \State $\mathbf{x}_i \gets \mathbf{x}_i + h\,\mathbf{v}_i$
               \Comment{inertial target $\mathbf{y}_i$}
    \EndFor
    \ForAll{constraints $j$}
        \State $\lambda_j \gets 0$ \Comment{multipliers reset every substep}
    \EndFor
    \For{$\mathrm{it} = 1$ \textbf{to} $n_{\mathrm{it}}$}
        \State \Call{ProjectConstraints}{$C_1,\dots,C_M,\ \mathbf{x},\ \boldsymbol{\lambda}$}
    \EndFor
    \ForAll{vertices $i$}
        \State $\mathbf{v}_i \gets (\mathbf{x}_i - \mathbf{x}^{\mathrm{prev}}_i)/h$
    \EndFor
\EndFor
\Statex
\Procedure{ProjectConstraints}{$C_1,\dots,C_M,\ \mathbf{x},\ \boldsymbol{\lambda}$}
    \ForAll{constraints $j$} \Comment{one Gauss--Seidel sweep}
        \State $\tilde{\alpha}_j \gets \alpha_j / h^2$
        \State $\Delta\lambda_j \gets \dfrac{-C_j(\mathbf{x}) - \tilde{\alpha}_j\,\lambda_j}
                {\nabla C_j\,\mathbf{M}^{-1}\nabla C_j^{T} + \tilde{\alpha}_j}$
        \State $\lambda_j \gets \lambda_j + \Delta\lambda_j$
        \State $\mathbf{x} \gets \mathbf{x} + \mathbf{M}^{-1}\nabla C_j^{T}\,\Delta\lambda_j$
    \EndFor
\EndProcedure
\end{algorithmic}
\end{algorithm}

\subsection{Characteristics}
\label{ssec:XPBDChar}
XPBD uses \emph{dual-variables}: an iteration's
state is encoded in the multipliers $\boldsymbol{\lambda}$, and positions are
derived from them. It handles \emph{stiffness through compliance}, since material
stiffness enters only via $\tilde{\alpha} = 1/(k h^{2})$, so very stiff
constraints naturally degenerate to hard projection ($\tilde\alpha \to 0$). It
uses \emph{first-order gradient information only}: each constraint contributes its
gradient $\nabla C_j$, and no Hessian of $E$ is built. It \emph{prefers
substeps over iterations}: its widely used small-steps optimization strategy only
uses one iteration but performs many substeps. As a member of the position-based
family it is also \emph{highly stable and computationally cheap}: it never
evaluates forces, working directly on positions through constraint projection
instead, which avoids the stiffness-driven instabilities of explicit force
integration and keeps the per-step cost low.

%----------------------------------------------------------------------------
\section{VBD: primal block-coordinate descent}
\label{sec:VBDAlgo}
%----------------------------------------------------------------------------

\subsection{Derivation}
\label{ssec:VBDDeriv}
VBD performs block-coordinate descent on $G$ in \eqref{eq:G}, with one block per
vertex. Holding all other vertices fixed, the update for vertex $i$ is the Newton
step on the local 3-vector problem
\begin{equation}
    \Delta\mathbf{x}_{i} \;=\; - \mathbf{H}_{i}^{-1}\,\mathbf{g}_{i},
\end{equation}
where
\begin{align}
    \mathbf{g}_{i} &= \frac{m_{i}}{h^{2}}\,(\mathbf{x}_{i} - \mathbf{y}_{i})
                      + \sum_{e \in \mathcal{F}_i} \frac{\partial E_{e}(\mathbf{x})}{\partial \mathbf{x}_i}, \\
    \mathbf{H}_{i} &= \frac{m_{i}}{h^{2}}\,\mathbf{I}_{3}
                      + \sum_{e \in \mathcal{F}_i} \frac{\partial^{2} E_{e}(\mathbf{x})}{\partial \mathbf{x}_i \mathbf{x}_i}.
\end{align}
Here $\mathbf{g}_{i}$ is the gradient of the local variational energy around
vertex $i$, $\mathcal{F}_i$ is the set of force elements acting on vertex $i$, and
$\mathbf{H}_{i}$ is the local Hessian. For a distance-spring energy
$E_{e} = \tfrac{1}{2} k\,(\|\mathbf{x}_{i} - \mathbf{x}_{j}\| - \ell_{0})^{2}$, the
per-vertex gradient and Hessian are closed-form and inexpensive to evaluate.

\subsection{Algorithm}
\label{ssec:VBDPseudo}
As with XPBD, the form given in Algorithm \autoref{alg:VBDCore} below is the simplified, core-comparison version: no
damping, no collisions, no Chebyshev acceleration. Adaptive initialization is kept on (defined in
\autoref{ssec:VBDAdaptiveInit}); the fairness rationale for both choices is given in
\autoref{ssec:DisableChebyshev}. Here I use the fixed-iteration-count variant in place of the
convergence-based early-stopping one. The full implemented version --- with damping, self-collision,
and Chebyshev acceleration --- is given in \autoref{sec:AppVBD} and uses the early-stopping variant.

\begin{algorithm}[!ht]
\caption{VBD step (core comparison). $n_{\mathrm{sub}}$ substeps, each with
         $n_{\mathrm{it}}$ block-descent iterations over the vertices.}
\label{alg:VBDCore}
\begin{algorithmic}[1]
\State $h \gets dt / n_{\mathrm{sub}}$
\For{$s = 1$ \textbf{to} $n_{\mathrm{sub}}$}
    \ForAll{vertices $i$}
        \State $\mathbf{a}^{\mathrm{ext}}_i \gets \mathbf{g} + w_i\,\mathbf{f}_{\mathrm{ext}}(\mathbf{x}_i)$
        \State $\mathbf{y}_i \gets \mathbf{x}_i + h\,\mathbf{v}_i + h^2\,\mathbf{a}^{\mathrm{ext}}_i$
               \Comment{inertial target}
        \State $\mathbf{x}^{\mathrm{prev}}_i \gets \mathbf{x}_i$
        \State $\mathbf{x}_i \gets \mathbf{x}_i + h\,\mathbf{v}_i + \tilde{a}_i\,h^2\,\mathbf{a}^{\mathrm{ext}}_i$
               \Comment{adaptive initial guess, $\tilde{a}_i \in [0,1]$}
    \EndFor
    \For{$\mathrm{it} = 1$ \textbf{to} $n_{\mathrm{it}}$}
        \State \Call{Sweep}{$\mathbf{x},\ \mathbf{y}$}
               \Comment{one Gauss--Seidel pass over vertices}
    \EndFor
    \ForAll{vertices $i$}
        \State $\mathbf{v}_i \gets (\mathbf{x}_i - \mathbf{x}^{\mathrm{prev}}_i)/h$
    \EndFor
\EndFor
\Statex
\Procedure{Sweep}{$\mathbf{x},\ \mathbf{y}$}
    \ForAll{vertices $i$}
        \State $\mathbf{g}_i \gets \dfrac{m_i}{h^2}(\mathbf{x}_i - \mathbf{y}_i)
                + \displaystyle\sum_{e \in \mathcal{F}_i} \dfrac{\partial E_e(\mathbf{x})}{\partial \mathbf{x}_i}$
               \Comment{gradient ($=-\mathbf{f}_i$)}
        \State $\mathbf{H}_i \gets \dfrac{m_i}{h^2}\mathbf{I}_3
                + \displaystyle\sum_{e \in \mathcal{F}_i} \dfrac{\partial^2 E_e(\mathbf{x})}{\partial \mathbf{x}_i \mathbf{x}_i}$
        \State $\mathbf{x}_i \gets \mathbf{x}_i - \mathbf{H}_i^{-1}\mathbf{g}_i$
               \Comment{local $3\times 3$ Newton step}
    \EndFor
\EndProcedure
\end{algorithmic}
\end{algorithm}

\subsection{Adaptive initialization}
\label{ssec:VBDAdaptiveInit}
The starting guess on line~5 of Algorithm \autoref{alg:VBDCore} blends two extremes per
vertex: the full inertial target $\mathbf{x}_i + h\mathbf{v}_i + h^{2}\mathbf{a}^{\mathrm{ext}}_i$
($\tilde{a}_i = 1$) and the pure velocity extrapolation
$\mathbf{x}_i + h\mathbf{v}_i$ ($\tilde{a}_i = 0$). The blend factor is read off
the previous step's acceleration
$\mathbf{a}^{\mathrm{prev}}_i = (\mathbf{v}_i^{t} - \mathbf{v}_i^{t-h})/h$ as
\begin{equation}
    \label{eq:VBDAdaptive}
    \tilde{a}_i \;=\; \mathrm{clamp}_{[0,1]}\!\left(
        \frac{\mathbf{a}^{\mathrm{prev}}_i \cdot \mathbf{a}^{\mathrm{ext}}_i}
             {\|\mathbf{a}^{\mathrm{ext}}_i\|^{2}}
    \right),
\end{equation}
which is the projection of the realized acceleration onto the external one.
$\tilde{a}_i \approx 1$ identifies a vertex that was in free fall last step,
so the inertial target is a good warm start. $\tilde{a}_i \approx 0$ identifies
a vertex that was held in place by stiff springs or a pin, where the inertial
target sits far from the minimizer of $G$ and the block-descent sweep would
otherwise spend iterations dragging it back; starting close to the previous
position is the better option there. XPBD has no analogous adaptive blend --- it
always starts from the full inertial target ($\tilde{a}=1$) and lets the
constraint sweeps do the rest. The reasoning for keeping adaptive initialization
enabled, and its acknowledged bias toward VBD, is
discussed in \autoref{ssec:DisableChebyshev}.

\subsection{Characteristics}
\label{ssec:VBDChar}
VBD is \emph{primal}: an
iteration's state is the positions $\mathbf{x}$, with no extra multipliers. It is
\emph{second-order locally}, since each block update is a Newton step on the true
elastic Hessian restricted to one vertex. It \emph{prefers iterations over
substeps}: VBD is normally used with one long step subdivided into many block
iterations, and the implicit Hessian handles stiffness even at larger $h$, thus
being \emph{stable even under large time steps}, which is especially valuable
when computational resources are constrained. Finally, it is \emph{parallel-friendly}: vertices that share
no constraint can be updated independently through graph coloring, which, when
parallelized, makes it particularly well suited to large-scale simulations.

%----------------------------------------------------------------------------
\section{Side-by-side: where the algorithms differ}
\label{sec:Sidebyside}
%----------------------------------------------------------------------------

\begin{table}[!ht]
    \footnotesize
    \centering
    \begin{tabular}{l l l}
        \toprule
                                  & XPBD                          & VBD \\
        \midrule
        Variable updated          & multipliers $\boldsymbol{\lambda}$ & positions $\mathbf{x}$ \\
        Order of information      & gradient $\nabla C$           & gradient + Hessian of $E$ \\
        Block size                & per-constraint                & per-vertex \\
        Stiffness handled by      & compliance $\tilde{\alpha}$   & implicit Hessian \\
        Typical configuration     & many substeps, $1$ iteration  & one step, many iterations \\
        Parallelism strategy      & constraint coloring (harder)  & vertex coloring (easier) \\
        \bottomrule
    \end{tabular}
    \caption{Algorithmic contrast between XPBD and VBD.}
    \label{tab:XPBDvsVBD}
\end{table}

The first and third rows capture the distinction that the primal/dual lens
(\autoref{sec:PrimalDualLens}) builds on, and the one the experiments return to
when interpreting each result.
