%----------------------------------------------------------------------------
\chapter{Experiments and results}
\label{chap:Experiments}
%----------------------------------------------------------------------------

This chapter reports measurements along the four evaluation axes from
\autoref{sec:EvalAxes}: efficiency, robustness, physical realism, and
flexibility. Each axis is designed to bear on at least one of the
hypotheses H1--H3 stated in \autoref{sec:Hypotheses}, and the synthesis is
deferred to \autoref{chap:Discussion}.

All experiments use the shared scaffolding and fairness controls described
in \autoref{sec:Fairness}: matched gravity, $dt = 1/24$~s, matched Rayleigh
damping, $\alpha_{\mathrm{XPBD}} = 1/k_{\mathrm{VBD}}$, Chebyshev and
adaptive initialisation disabled in VBD for the core comparison.

%----------------------------------------------------------------------------
\section{Computational efficiency --- cloth}
\label{sec:EffResults}
%----------------------------------------------------------------------------

\subsection{Frame time vs.\ node count}
\label{ssec:MsVsNodes}
I measure isolated solver time (ms/frame) over a hanging-cloth scene at
mesh resolutions \emph{TODO: list resolutions, e.g.\ $10\times 10$ to
$100\times 100$}, holding solver settings fixed at the defaults of
\autoref{sec:Fairness}.

\emph{TODO: Figure --- ms/frame vs.\ node count, XPBD and VBD curves.}

\subsection{Frame time vs.\ substep / iteration count}
\label{ssec:MsVsCount}
For a fixed mesh resolution I sweep:
\begin{itemize}
    \item XPBD substep count $n_{\mathrm{sub}} \in \{1, 2, 4, 8, 16, 32\}$
    with a single inner iteration (the small-steps configuration),
    \item VBD iteration count $n_{\mathrm{it}} \in \{1, 2, 4, 8, 16, 32\}$.
\end{itemize}

\emph{TODO: Figure --- ms/frame vs.\ work count, both solvers on common
axis.}

\subsection{Maximum node counts at 30 and 60~fps}
\label{ssec:MaxNodes}
Following Ma et~al.~\cite{app142311072}, I report the largest mesh
sustaining 30~fps and 60~fps for each solver at default
settings.

\emph{TODO: Table --- max node count, two thresholds, two solvers.}

\subsection{Convergence: residual vs.\ work}
\label{ssec:ResidualConvergence}
For a moderately stretched cloth in isolation, I measure the variational
residual $\|\nabla G(\mathbf{x})\|$ as a function of iteration count and as
a function of wall-clock time, plotting both solvers against a Newton
reference. This is the analogue of Fig.~17 in \cite{Chen2024}.

\emph{TODO: Figure --- residual vs.\ iterations and residual vs.\ wall-clock.}

\subsection{Equal-wall-clock quality frontier}
\label{ssec:EqualWallClock}
The headline efficiency result. I compute the solution error (vs.\ Newton)
at every combination of $n_{\mathrm{sub}}$ for XPBD and $n_{\mathrm{it}}$
for VBD, and plot error against ms/frame on a log--log axis. The lower
envelope of the two clouds is each solver's quality frontier; the
\emph{equal-wall-clock} comparison reads off which solver achieves lower
error at a given ms/frame budget.

\emph{TODO: Figure --- error vs.\ ms/frame Pareto fronts.}

\subsection{VBD with Chebyshev and adaptive initialisation (production mode)}
\label{ssec:VBDChebyshev}
For completeness I re-run the equal-wall-clock experiment of
\autoref{ssec:EqualWallClock} with VBD's Chebyshev acceleration and
adaptive initialisation re-enabled. This is the configuration a
practitioner would deploy; the result is reported separately because, per
\autoref{ssec:DisableChebyshev}, it does not isolate the primal/dual
distinction --- it measures VBD plus its accelerations against vanilla XPBD.

\emph{TODO: Figure / table --- accelerated VBD on the same axes as
\autoref{ssec:EqualWallClock}.}

%----------------------------------------------------------------------------
\section{Robustness and stability}
\label{sec:Robustness}
%----------------------------------------------------------------------------

\subsection{Stability map: max stable $dt$ vs.\ stiffness}
\label{ssec:StabilityMap}
For each solver I sweep $dt$ and material stiffness $k$, flagging
combinations where artefacts appear (oscillation amplitude exceeds rest
length, NaNs, visible blow-up). The deliverable is a 2D stability map per
solver, summarising at a glance the operating envelope of each. This
replaces the simpler ``raise $dt$ until artefacts appear'' test of
\cite{app142311072} (their Figs.~7--8) with a fuller picture.

\emph{TODO: Figure --- 2D stability map for XPBD and VBD on shared axes.}

\subsection{Single-iteration regime}
\label{ssec:SingleIter}
Both solvers are configured to perform exactly one inner iteration per
substep, and I report whether each remains visually plausible across the
default scenes. This is the most adversarial setting for each algorithm
and isolates per-iteration update quality.

\emph{TODO: Side-by-side renders, qualitative description.}

\subsection{High mass ratio at the end of a pendulum}
\label{ssec:MassRatioPendulum}
A single spring pendulum with a heavy weight at the free end and a light
chain leading to the pivot. I sweep the mass ratio $m_{\mathrm{tip}} /
m_{\mathrm{chain}}$ from $1$ to $10^{4}$ and report each solver's
qualitative behaviour and quantitative position error vs.\ Newton.
\emph{Hypothesis H1 prediction: XPBD degrades.}

\emph{TODO: Figure --- error vs.\ mass ratio.}

\subsection{High mass ratio chain}
\label{ssec:MassRatioChain}
A chain with alternating heavy and light particles, designed to maximise
the chain length over which mass disparities couple. Same outputs as the
pendulum.

\emph{TODO: Figure / description.}

\subsection{High stiffness ratio chain}
\label{ssec:StiffnessRatioChain}
A chain with alternating stiff and soft springs. Sweep the stiffness ratio
from $1$ to $10^{4}$. \emph{Hypothesis H2 prediction: VBD degrades.}

\emph{TODO: Figure --- error vs.\ stiffness ratio.}

\subsection{Self-collision: cloth folded onto itself}
\label{ssec:SelfCollResults}
A cloth dropped onto a small obstacle, folding onto itself. Qualitative
visual comparison, noting that the contact model differs (projection vs.\
penalty) per \autoref{sec:SelfCollision}.

\emph{TODO: Renders, qualitative description.}

%----------------------------------------------------------------------------
\section{Physical realism}
\label{sec:Realism}
%----------------------------------------------------------------------------

\subsection{Newton reference}
\label{ssec:NewtonRef}
I build a high-iteration Newton solver targeting $\|\nabla G\| < 10^{-9}$
(or until further iterations reduce $G$ by less than $10^{-12}$). The
Newton solver shares storage and energies with the production solvers, so
the comparison is over algorithm, not material. The error metric throughout
is the mass-weighted position error
$\|\mathbf{x} - \mathbf{x}^{\star}\|_{\mathbf{M}}$, where
$\mathbf{x}^{\star}$ is the Newton solution.

\subsection{Harmonic oscillator}
\label{ssec:Harmonic}
A single spring-mass system oscillating in 1D, compared against the
analytic solution $x(t) = A\cos(\omega t - \phi)\,e^{-\zeta\omega t}$.
I measure:
\begin{itemize}
    \item period as a function of $h$ and of iteration / substep count,
    \item amplitude decay rate as a function of damping coefficient.
\end{itemize}
This is the analogue of Fig.~2 in \cite{10.1145/2994258.2994272}.

\emph{TODO: Figure --- measured period and decay, both solvers, against
analytic.}

\subsection{Energy tracking}
\label{ssec:EnergyTracking}
For an undamped free-falling and bouncing cloth I plot total mechanical
energy over time, identifying systematic energy gain (instability) or loss
(numerical damping) for each solver.

\emph{TODO: Figure --- energy vs.\ time, both solvers.}

\subsection{Hanging cloth: stretched area vs.\ reference}
\label{ssec:HangingCloth}
A square cloth pinned at the top corners, allowed to settle. I measure
the steady-state stretched area against the Newton reference. This is a
realism check at static equilibrium, where transient dynamics drop out.

\emph{TODO: Figure --- stretched area error vs.\ stiffness.}

\subsection{Damping cross-check}
\label{ssec:DampingCheck}
At matched Rayleigh coefficients $\alpha_{\mathrm{Rayleigh}}$ and
$\beta_{\mathrm{Rayleigh}}$, I measure the amplitude-decay curve produced
by each solver on the harmonic oscillator. Identical decay confirms the
damping models of \autoref{sec:Damping} are calibrated consistently; this
both validates fairness and is a result in its own right.

\emph{TODO: Figure --- decay envelopes overlaid.}

%----------------------------------------------------------------------------
\section{Flexibility}
\label{sec:FlexResults}
%----------------------------------------------------------------------------

\subsection{Iteration-count independence of stiffness}
\label{ssec:IterIndepStiff}
For each solver I sweep stiffness $k$ over several decades and measure how
many iterations / substeps are needed to reach a fixed error threshold.
Following Fig.~6 in \cite{10.1145/2994258.2994272}, XPBD is expected to be
roughly stiffness-independent in iteration count thanks to the compliance
formulation.

\emph{TODO: Figure --- iterations to threshold vs.\ stiffness.}

\subsection{Stiffness-ratio test}
\label{ssec:StiffRatioTest}
The same kind of plot as the previous subsection, but varying the
\emph{ratio} of stiffnesses across the mesh rather than a uniform stiffness.
This is the experiment of \cite{Chen2024} Fig.~24 (and AVBD Fig.~4). I
include a brief discussion of AVBD's improvement on this axis without
implementing it myself.

\emph{TODO: Figure --- error vs.\ stiffness ratio, with AVBD reference
qualitatively cited.}

\subsection{Ease of adding a new constraint or force}
\label{ssec:NewConstraint}
I implement a representative additional force --- \emph{TODO: choose, e.g.\
bending-angle constraint or area-preservation force} --- in both solvers
and report:
\begin{itemize}
    \item lines of code added,
    \item whether a Hessian derivation was required (VBD: yes; XPBD: no, a
    gradient suffices),
    \item time to first correct simulation.
\end{itemize}
The metric is admittedly subjective but captures a real practitioner cost
that the more quantitative experiments above do not.

\emph{TODO: Table --- LOC, derivation effort, time-to-correct.}
