%----------------------------------------------------------------------------
\chapter{\bevezetes}
\label{chap:Introduction}
%----------------------------------------------------------------------------

%----------------------------------------------------------------------------
\section{Motivation}
\label{sec:Motivation}
%----------------------------------------------------------------------------
Real-time simulation has become a baseline expectation in games, virtual
reality, and augmented reality. A simulator must deliver believable dynamics
within a tight frame budget: at 60~fps, only about 16~ms is available for all
physics, rendering, and gameplay logic combined, and physics receives only a
fraction of that.

This thesis focuses on deformable bodies --- springs, chains, and cloth ---
rather than rigid bodies, for several reasons. Deformable simulation is the
harder case, with many more degrees of freedom and constraints per frame, so it
stresses both solvers more heavily. Even when vertices and constraints number in
the tens of thousands, every part of the simulated system remains visible, so
solver behavior can be observed directly and failures are easy to spot. Springs,
moreover, are among the most common and most easily measurable primitives in
simulation, while cloth is widely used in practice and a crucial stepping
stone toward full soft-body simulation.

Designing a real-time simulator of this kind is hard for three recurring
reasons. The first is \emph{stiffness}: cloth and soft tissue impose very stiff
constraints --- almost-inextensible fabric, near-rigid bone --- and stiff systems
make explicit integrators unstable unless the time step shrinks to impractically
small values. The second is \emph{stability}: even implicit integrators can blow
up under poorly conditioned constraints, large mass ratios between connected
particles, or the velocity discontinuities introduced by contact. The third is
the \emph{strict time constraint}: methods that converge well offline, such as
Newton's method or projective dynamics, are too expensive per frame, while cheap
real-time methods typically trade convergence quality for predictable cost.

\paragraph{The accuracy/efficiency trade-off.}
Taken together, these pressures place every physics-based simulator somewhere on
a line between \emph{physical accuracy} (combining stiffness and stability) and
\emph{computational efficiency}; pushing toward efficiency inevitably costs
realism. \autoref{tab:accuracy-efficiency} sketches this spectrum for the
best-known deformable-body solvers, ordered from most accurate to most efficient.

At the efficient extreme, the Mass-Spring System (MSS) represents a deformable
material as a network of point masses connected by springs and advances it with
explicit Euler integration --- the most intuitive and straightforward approach.
Although it is in principle more physically grounded than PBD or XPBD, it is
highly unstable for stiff and complex systems, which pushes it to the bottom of
the list: explicit integration of stiff cloth requires impractically small time
steps to stay stable, which is exactly what motivated the move to implicit
integration~\cite{10.1145/280814.280821}. At the accurate extreme,
the Finite Element Method (FEM) treats an object as a continuous material
discretized into volumetric finite elements (such as tetrahedra) and solves the
resulting implicit-Euler problem with a Newton solver. It excels at physical
accuracy, but its high computational cost makes it impractical for real-time
applications.~\cite{nikishkov2004introfem}

\begin{table}[h]
\centering
\begin{tabular}{l l l}
\toprule
\textbf{Method} & \textbf{Accuracy} & \textbf{Cost} \\
\midrule
FEM     & reference            & most expensive \\
VBD     & moderate--high       & moderate       \\
XPBD    & moderate             & cheap          \\
PBD     & low (stiffness-dep.) & cheap          \\
MSS     & poor (unstable)      & cheapest       \\
\bottomrule
\end{tabular}
\caption{Deformable-simulation methods ordered from most accurate to most
efficient. XPBD and VBD are in the real-time domain.}
\label{tab:accuracy-efficiency}
\end{table}

XPBD and VBD both sit near the efficient end of this spectrum, inside the
real-time category. XPBD~\cite{10.1145/2994258.2994272} has been a workhorse for
years, whereas VBD~\cite{Chen2024} is a recent and promising addition that has
not yet been thoroughly evaluated head-to-head against XPBD.

%----------------------------------------------------------------------------
\section{Contribution}
\label{sec:Contribution}
%----------------------------------------------------------------------------
The central contribution of this thesis is, to the best of my knowledge, the
\emph{first faithful XPBD-vs-VBD comparison}. Both algorithms are implemented in
the same Unity-based codebase, applied to the same mass-spring material model,
and run under a matched compute budget --- matched iteration and substep counts,
not merely matched material parameters. They are evaluated not only qualitatively
but also against a high-iteration Newton solver that serves as a numerical
ground-truth reference, which lets me report \emph{solution error} rather than
only relative performance.

Crucially, the comparison uses XPBD --- the constraint-stiffness-independent
variant that any practitioner would deploy today --- together with the canonical
VBD formulation, rather than plain PBD or a non-canonical variant. This sharpens
the closest prior comparison, Ma et~al.~\cite{app142311072}, which contrasts
plain PBD with a non-canonical VBD implementation and reports only frame-rate
scaling; \autoref{sec:Ma2024} sets out exactly how my setup differs.

%----------------------------------------------------------------------------
\section{The primal/dual lens}
\label{sec:PrimalDualLens}
%----------------------------------------------------------------------------
The two solvers sit on opposite sides of the \emph{primal/dual} duality
identified by Macklin et~al.~\cite{10.1111/cgf.14104}: XPBD is the dual solver,
solving for Lagrange multipliers on the constraints, while VBD is the primal
solver, solving directly for vertex positions. This duality is the organizing
lens of the thesis: it predicts complementary weak spots --- one per solver ---
that the stress tests in \autoref{chap:Experiments} are built around.
\autoref{sec:PrimalDual} derives the distinction and the failure-mode
predictions in full.

%----------------------------------------------------------------------------
\section{Evaluation axes}
\label{sec:EvalAxes}
%----------------------------------------------------------------------------
I evaluate both solvers along four axes that together capture the trade-offs a
practitioner cares about:
\begin{description}
    \item[Efficiency.] Computational cost per frame as a function of mesh size
        and substep/iteration count.
    \item[Robustness.] Behavior under stress: extreme mass ratios, extreme
        stiffness ratios, and the stability issues they produce.
    \item[Physical realism.] Energy behavior, damping behavior, and how close
        each method comes to a ground-truth Newton reference.
    \item[Flexibility.] The independence of material properties from iteration
        count, the ease of adding new constraint types, and the domain coverage
        of each model.
\end{description}

%----------------------------------------------------------------------------
\section{Thesis structure}
\label{sec:Structure}
%----------------------------------------------------------------------------
The remainder of the thesis is organized as follows.
\autoref{chap:Background} lays the theoretical foundation: the explicit and
implicit time-stepping schemes, the variational form of implicit Euler that
unifies both solvers as approximate minimizers of the same energy, the
mass-spring model, the Gauss--Seidel structure shared by both solvers, and the
primal/dual distinction that organizes the comparison.
\autoref{chap:RelatedWork} traces the PBD-to-XPBD and VBD lineages and positions
this work against the closest prior comparison, Ma et~al.~\cite{app142311072}.
\autoref{chap:Algorithms} states both algorithms in their canonical
core-comparison form --- derivation, pseudocode, and characteristics --- and
contrasts them side by side. \autoref{chap:Methodology} describes the Unity
implementation, the cloth model, and the fairness controls that keep the
comparison balanced, along with self-collision handling and the
parallelization outlook. \autoref{chap:Experiments} presents the results along
the four evaluation axes, measured against a Newton reference.
\autoref{chap:Discussion} synthesizes those results, gives a verdict on each
axis, and sets out the threats to validity. \autoref{chap:Conclusion} closes
with a summary and directions for future work.
