%----------------------------------------------------------------------------
\chapter{Related work}
\label{chap:RelatedWork}
%----------------------------------------------------------------------------

This chapter introduces the position-based and primal-descent lineages whose
endpoints --- XPBD and VBD --- this thesis compares, and positions my work
relative to the closest prior comparison.

%----------------------------------------------------------------------------
\section{From PBD to XPBD}
\label{sec:PBDtoXPBD}
%----------------------------------------------------------------------------
Position-Based Dynamics (PBD), introduced by M\"uller et~al.~\cite{MULLER2007109},
targets real-time cloth and soft-body simulation in games. Each iteration
projects the predicted positions onto the constraint set --- moving the particles
each constraint couples until the constraint is satisfied --- through
Gauss--Seidel sweeps, and recovers velocities by finite difference. PBD is unconditionally stable, but it has a
well-known flaw: \emph{constraint stiffness depends on the iteration count and the
time step}.

XPBD, introduced by Macklin et~al.~\cite{10.1145/2994258.2994272}, fixes this by
augmenting each constraint with a compliance term $\alpha = 1/k$ and solving for
the associated Lagrange multiplier $\lambda$. The resulting iteration solves an
implicit-Euler problem with arbitrary elastic energies, and constraint stiffness
becomes a true material parameter, independent of the iteration count. As Ma
et~al.~\cite{app142311072} summarize the lineage:

\begin{quote}
``Early research focused on force--object relationships, leading to the
development of models such as the Mass-Spring System, Finite Element Method,
and Rigid Body Dynamics. However, these models have limitations,
including performance degradation, lack of real-time capability, and unrealistic
motion as simulation complexity increases. PBD was introduced to address these
limitations. Initially developed by Jakobsen, PBD provides a stable and
efficient approach by applying wedge-based constraints instead of force-based
methods. Macklin et al.\ further refined and optimized PBD for real-time
interaction.''
\end{quote}

\paragraph{Small steps in physics simulation.}
Macklin et~al.~\cite{Macklin2019SmallSI} (the ``Small Steps'' note) later argued
that XPBD converges most reliably with \emph{many short substeps and a single
iteration per substep}, rather than a single long step with many iterations. With only one iteration per
substep the multiplier term entering the update is always zero --- $\lambda$ is
reset each substep and never re-used across iterations --- so there is no need to
store and carry it between iterations. This version is now a standard XPBD configuration, but
for the sake of a fair comparison --- to compare methods across substep and iteration counts --- I deliberately leave this optimization out of
the core comparison.

%----------------------------------------------------------------------------
\section{Vertex Block Descent (VBD)}
\label{sec:VBDLineage}
%----------------------------------------------------------------------------
Vertex Block Descent, due to Chen et~al.~\cite{Chen2024}, treats each vertex as a
tiny optimization block: at each iteration, every vertex moves to the minimizer of
$G$ with all other vertices held fixed. This requires the local $3\times 3$
Hessian of the energies sharing the vertex and a Newton step on that block. Ma
et~al.~\cite{app142311072} place the method as follows:

\begin{quote}
``More recently, the VBD algorithm that has been proposed provides an efficient
approach to parallelize large-scale computations. By processing nodes in block
units, VBD significantly enhances performance, particularly in simulations
involving a high number of nodes.''
\end{quote}

The original VBD paper contributes three features beyond the basic block descent.
\emph{Chebyshev acceleration} is an over-relaxation that significantly speeds
convergence, at the cost of the method's energy guarantees. \emph{Adaptive
initialization} seeds each step's initial guess from a blend of the full inertial
prediction and an inertia-only extrapolation, leaning on the latter for vertices
near rest to avoid over-shooting under gravity; the exact blend formula is given
in \autoref{ssec:VBDAdaptiveInit}. \emph{Graph-coloring parallelism} lets
vertices that share no constraint be updated independently, enabling massive
parallelism. The fairness rationale for enabling adaptive initialization while
disabling Chebyshev in the core comparison is given in
\autoref{ssec:DisableChebyshev}.

A follow-up, Augmented VBD (AVBD)~\cite{Giles2025}, improves stiffness-ratio
handling, infinite-stiffness handling, and convergence behavior. To keep the
baseline comparison anchored to the established formulation, I restrict the
implementation to the canonical 2024 VBD; I cite AVBD in
\autoref{sec:FlexResults} as the most relevant counter-evidence on the
stiffness-ratio axis, and revisit the question it raises in
\autoref{ssec:ThreatAVBD}.

%----------------------------------------------------------------------------
\section{Prior comparison: Ma et~al.\ (2024)}
\label{sec:Ma2024}
%----------------------------------------------------------------------------
The most directly comparable prior work is Ma et~al.~\cite{app142311072}, who
compare cloth simulation under PBD and VBD in Unity and establish that VBD scales
better than PBD with node count. My work differs in four substantial ways:
\begin{enumerate}
    \item \textbf{XPBD, not PBD.} Plain PBD has the stiffness-coupling flaw that
        XPBD was designed to fix, so a PBD-vs-VBD comparison conflates algorithmic
        differences with this known bug. Using XPBD isolates the primal/dual
        distinction.
    \item \textbf{Canonical VBD.} Ma et~al.'s pseudocode departs from
        \cite{Chen2024} in two concrete ways relevant to the comparison: it does
        not include the adaptive-initialization blend
        (\autoref{ssec:VBDAdaptiveInit}), which always starts each step from the
        full inertial target; and it does not include Chebyshev acceleration.
        Both omissions affect convergence on stiff problems, where the canonical
        warm start and over-relaxation carry most of VBD's reported per-step
        speedup. I follow canonical VBD, but disable Chebyshev for the core
        comparison so I read off algorithmic properties rather than acceleration
        tricks; adaptive initialization is kept enabled because it is VBD's
        inertial prediction and has a direct counterpart in XPBD's explicit
        guess. The fairness rationale is detailed in
        \autoref{ssec:DisableChebyshev}.
    \item \textbf{Beyond frame rate.} Ma et~al.\ report fps and node counts. I add
        a Newton ground truth, a solution quality vs.\ computational cost comparison,
        and targeted stress tests (mass ratio, stiffness ratio, large $dt$) keyed to the
        primal/dual distinction.
    \item \textbf{Varying the substep count.} Their comparison kept the substep
        count fixed throughout, so it did not consider how changing it affects each
        algorithm.
\end{enumerate}

In short, this thesis answers a question that Ma et~al.\ explicitly left as future
work: how does VBD compare to \emph{XPBD}, and at \emph{matched quality} rather
than only at matched node counts?

%----------------------------------------------------------------------------
\section{Reference materials}
\label{sec:RefMaterials}
%----------------------------------------------------------------------------
The Physics-Based Simulation book of Li et~al.~\cite{li2026physics} provided
canonical derivations and pedagogical notes; its treatment of implicit Euler as
variational minimization directly informs the framing in \autoref{sec:Variational}.
The \emph{Ten Minute Physics} tutorial series~\cite{tenMinutePhysics} was used as
a sanity-check reference for the correctness of the XPBD implementation, and the
TinyVBD reference implementation~\cite{vbdGithub} served the same role for the
VBD side.
